Lost in the woods: shifting habitats can lead phylogeography astray

### 2025-05-17 manuscript/phylogeo.tex
inverse affiliation
inline bib

\begin{document}
%\title{On the robustness and consistency of continuous phylogeographic inference}
%\title{Phylogeography: unsafe at any speed}
\title{Lost in the woods: shifting habitats can lead phylogeography astray}
\author{Richard A.~Neher}
\affiliation{Swiss Institute of Bioinformatics, and Biozentrum, University of Basel, Switzerland \looseness=-5}

@@

\begin{figure}[tb]
    \includegraphics*[width=\columnwidth]{figures/illustration_tree.pdf}
    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees with $N=100$ tips and the spatial location of the tips.}
    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present.
    Temporal and spatial scales are determined by parameter choice, only tree shape and spatial patterns are meaningful. }
\end{figure}

@@

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

\begin{figure}[!h]
    \includegraphics[width=0.48\textwidth]{figures/stable_density}
    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat (dashed horizontal line). The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
    For $r\ll L$, density fluctuations are strongly suppressed.
    When density regulation homogenizes density, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
    The dashed vertical line indicates $D=L^2/2T_c$.}
\end{figure}

@@

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/seasaw_example}
    \includegraphics[width=\textwidth]{figures/seasaw}
    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows the time course of the carrying capacity through half a period. The color indicates time along the periodic habitat oscillations with yellow and purple representing most turning points of the oscillations. The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back in the second half of the period. The habitat size is $L_x=3$, $L_y=1$.
    {\bf Panel B} shows a tree from a simulation under these conditions with $T=500$, $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
    The population is sampled at a time when the carrying capacity is flat after having moved to the left, that is the population is about to spread to the right.
    {\bf Panel D} show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
    {\bf Panel E} shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes. Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$ (data for $z_y^2$ are hard to see in the figure as $z_y^2$ is of order 1 as expected).
    The summary inferences in panels D \& E are made from many trees each of which represents samples from a single time point.}
\end{figure*}

@@

\begin{figure}[tb]
    \includegraphics[width=0.3\textwidth]{figures/traveling_wave}
    \includegraphics[width=0.48\textwidth]{figures/waves}
    \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
    {\bf Panel A} shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
    {\bf Panels B\&C} show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
\end{figure}

@@

\bibliography{bib}

>>>>

\begin{document}
\title{Lost in the woods: shifting habitats can lead phylogeography astray}
\author{Richard A.~Neher}
\affiliation{Biozentrum, University of Basel and Swiss Institute of Bioinformatics, Switzerland \looseness=-5}
\email{richard.neher@unibas.ch}

@@

\begin{figure}[tb]
    \includegraphics*[width=\columnwidth]{figures/Fig1}
    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees with $N=100$ tips and the spatial location of the tips.}
    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present.
    Temporal and spatial scales are determined by parameter choice, only tree shape and spatial patterns are meaningful. }
\end{figure}

@@

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/Fig2}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

\begin{figure}[!h]
    \includegraphics[width=0.48\textwidth]{figures/Fig3}
    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat (dashed horizontal line). The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
    For $r\ll L$, density fluctuations are strongly suppressed.
    When density regulation homogenizes density, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
    The dashed vertical line indicates $D=L^2/2T_c$.}
\end{figure}

@@

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/Fig4}
    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows the time course of the carrying capacity through half a period. The color indicates time along the periodic habitat oscillations with yellow and purple representing most turning points of the oscillations. The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back in the second half of the period. The habitat size is $L_x=3$, $L_y=1$.
    {\bf Panel B} shows a tree from a simulation under these conditions with $T=500$, $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
    The population is sampled at a time when the carrying capacity is flat after having moved to the left, that is the population is about to spread to the right.
    {\bf Panel D} show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
    {\bf Panel E} shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes. Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$ (data for $z_y^2$ are hard to see in the figure as $z_y^2$ is of order 1 as expected).
    The summary inferences in panels D \& E are made from many trees each of which represents samples from a single time point.}
\end{figure*}

@@

\begin{figure}[tb]
    \includegraphics[width=\columnwidth]{figures/Fig5}
    \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
    {\bf Panel A} shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
    {\bf Panels B\&C} show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
\end{figure}

@@

\begin{thebibliography}{27}
    \expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
    \expandafter\ifx\csname bibnamefont\endcsname\relax
      \def\bibnamefont#1{#1}\fi
    \expandafter\ifx\csname bibfnamefont\endcsname\relax
      \def\bibfnamefont#1{#1}\fi
    \expandafter\ifx\csname citenamefont\endcsname\relax
      \def\citenamefont#1{#1}\fi
    \expandafter\ifx\csname url\endcsname\relax
      \def\url#1{\texttt{#1}}\fi
    \expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
    \providecommand{\bibinfo}[2]{#2}
    \providecommand{\eprint}[2][]{\url{#2}}

    \bibitem[{\citenamefont{Bolker and Pacala}(1997)}]{bolker_using_1997}
    \bibinfo{author}{\bibnamefont{Bolker}, \bibfnamefont{B.}}, and
      \bibinfo{author}{\bibfnamefont{S.~W.} \bibnamefont{Pacala}},
      \bibinfo{year}{1997}, \bibinfo{journal}{Theoretical Population Biology}
      \textbf{\bibinfo{volume}{52}}(\bibinfo{number}{3}), \bibinfo{pages}{179},
      ISSN \bibinfo{issn}{0040-5809}.

    \bibitem[{\citenamefont{Brunker} \emph{et~al.}(2015)\citenamefont{Brunker,
      Marston, Horton, Cleaveland, Fooks, Kazwala, Ngeleja, Lembo, Sambo, Mtema,
      Sikana, Wilkie} \emph{et~al.}}]{brunker_elucidating_2015}
    \bibinfo{author}{\bibnamefont{Brunker}, \bibfnamefont{K.}},
      \bibinfo{author}{\bibfnamefont{D.~A.} \bibnamefont{Marston}},
      \bibinfo{author}{\bibfnamefont{D.~L.} \bibnamefont{Horton}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Cleaveland}},
      \bibinfo{author}{\bibfnamefont{A.~R.} \bibnamefont{Fooks}},
      \bibinfo{author}{\bibfnamefont{R.}~\bibnamefont{Kazwala}},
      \bibinfo{author}{\bibfnamefont{C.}~\bibnamefont{Ngeleja}},
      \bibinfo{author}{\bibfnamefont{T.}~\bibnamefont{Lembo}},
      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Sambo}},
      \bibinfo{author}{\bibfnamefont{Z.~J.} \bibnamefont{Mtema}},
      \bibinfo{author}{\bibfnamefont{L.}~\bibnamefont{Sikana}},
      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Wilkie}}, \emph{et~al.},
      \bibinfo{year}{2015}, \bibinfo{journal}{Virus Evolution}
      \textbf{\bibinfo{volume}{1}}(\bibinfo{number}{1}), \bibinfo{pages}{vev011},
      ISSN \bibinfo{issn}{2057-1577}.

    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2024)\citenamefont{Dellicour,
      Bastide, Rocu, Fargette, Hardy, Suchard, Guindon, and
      Lemey}}]{dellicour_how_2024}
    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Bastide}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Rocu}},
      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Fargette}},
      \bibinfo{author}{\bibfnamefont{O.~J.} \bibnamefont{Hardy}},
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Guindon}}, and
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
      \bibinfo{year}{2024}, \bibinfo{journal}{PLOS Biology}
      \textbf{\bibinfo{volume}{22}}(\bibinfo{number}{12}),
      \bibinfo{pages}{e3002914}, ISSN \bibinfo{issn}{1545-7885},
      \bibinfo{note}{publisher: Public Library of Science}.

    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2021)\citenamefont{Dellicour,
      Gill, Faria, Rambaut, Pybus, Suchard, and Lemey}}]{dellicour_relax_2021}
    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
      \bibinfo{author}{\bibfnamefont{M.~S.} \bibnamefont{Gill}},
      \bibinfo{author}{\bibfnamefont{N.~R.} \bibnamefont{Faria}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
      \bibinfo{author}{\bibfnamefont{O.~G.} \bibnamefont{Pybus}},
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}}, and
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
      \bibinfo{year}{2021}, \bibinfo{journal}{Molecular Biology and Evolution}
      \textbf{\bibinfo{volume}{38}}(\bibinfo{number}{8}), \bibinfo{pages}{3486},
      ISSN \bibinfo{issn}{1537-1719}.

    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2020)\citenamefont{Dellicour,
      Lequime, Vrancken, Gill, Bastide, Gangavarapu, Matteson, Tan, du~Plessis,
      Fisher, Nelson, Gilbert} \emph{et~al.}}]{dellicour_epidemiological_2020}
    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Lequime}},
      \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Vrancken}},
      \bibinfo{author}{\bibfnamefont{M.~S.} \bibnamefont{Gill}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Bastide}},
      \bibinfo{author}{\bibfnamefont{K.}~\bibnamefont{Gangavarapu}},
      \bibinfo{author}{\bibfnamefont{N.~L.} \bibnamefont{Matteson}},
      \bibinfo{author}{\bibfnamefont{Y.}~\bibnamefont{Tan}},
      \bibinfo{author}{\bibfnamefont{L.}~\bibnamefont{du~Plessis}},
      \bibinfo{author}{\bibfnamefont{A.~A.} \bibnamefont{Fisher}},
      \bibinfo{author}{\bibfnamefont{M.~I.} \bibnamefont{Nelson}},
      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}}, \emph{et~al.},
      \bibinfo{year}{2020}, \bibinfo{journal}{Nature Communications}
      \textbf{\bibinfo{volume}{11}}(\bibinfo{number}{1}), \bibinfo{pages}{5620},
      ISSN \bibinfo{issn}{2041-1723}, \bibinfo{note}{publisher: Nature Publishing
      Group}.

    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2017)\citenamefont{Dellicour,
      Rose, Faria, Vieira, Bourhy, Gilbert, Lemey, and
      Pybus}}]{dellicour_using_2017}
    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
      \bibinfo{author}{\bibfnamefont{R.}~\bibnamefont{Rose}},
      \bibinfo{author}{\bibfnamefont{N.~R.} \bibnamefont{Faria}},
      \bibinfo{author}{\bibfnamefont{L.~F.~P.} \bibnamefont{Vieira}},
      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Bourhy}},
      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}}, and
      \bibinfo{author}{\bibfnamefont{O.~G.} \bibnamefont{Pybus}},
      \bibinfo{year}{2017}, \bibinfo{journal}{Molecular Biology and Evolution}
      \textbf{\bibinfo{volume}{34}}(\bibinfo{number}{10}), \bibinfo{pages}{2563},
      ISSN \bibinfo{issn}{0737-4038}.

    \bibitem[{\citenamefont{Desai and Nelson}(2005)}]{desai_quasispecies_2005}
    \bibinfo{author}{\bibnamefont{Desai}, \bibfnamefont{M.~M.}}, and
      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
      \bibinfo{year}{2005}, \bibinfo{journal}{Theoretical Population Biology}
      \textbf{\bibinfo{volume}{67}}(\bibinfo{number}{1}), \bibinfo{pages}{33}, ISSN
      \bibinfo{issn}{0040-5809}.

    \bibitem[{\citenamefont{Etheridge}(2004)}]{etheridge_survival_2004}
    \bibinfo{author}{\bibnamefont{Etheridge}, \bibfnamefont{A.~M.}},
      \bibinfo{year}{2004}, \bibinfo{journal}{The Annals of Applied Probability}
      \textbf{\bibinfo{volume}{14}}(\bibinfo{number}{1}), \bibinfo{pages}{188},
      ISSN \bibinfo{issn}{1050-5164}, \bibinfo{note}{publisher: Institute of
      Mathematical Statistics}.

    \bibitem[{\citenamefont{Felsenstein}(1975)}]{felsenstein_pain_1975}
    \bibinfo{author}{\bibnamefont{Felsenstein}, \bibfnamefont{J.}},
      \bibinfo{year}{1975}, \bibinfo{journal}{The American Naturalist}
      \textbf{\bibinfo{volume}{109}}(\bibinfo{number}{967}), \bibinfo{pages}{359},
      ISSN \bibinfo{issn}{0003-0147}, \bibinfo{note}{publisher: The University of
      Chicago Press}.

    \bibitem[{\citenamefont{Fisher}(1937)}]{fisher_wave_1937}
    \bibinfo{author}{\bibnamefont{Fisher}, \bibfnamefont{R.~A.}},
      \bibinfo{year}{1937}, \bibinfo{journal}{Annals of Eugenics}
      \textbf{\bibinfo{volume}{7}}(\bibinfo{number}{4}), \bibinfo{pages}{355}, ISSN
      \bibinfo{issn}{2050-1439}.

    \bibitem[{\citenamefont{Gill} \emph{et~al.}(2017)\citenamefont{Gill, Tung~Ho,
      Baele, Lemey, and Suchard}}]{gill_relaxed_2017}
    \bibinfo{author}{\bibnamefont{Gill}, \bibfnamefont{M.~S.}},
      \bibinfo{author}{\bibfnamefont{L.~S.} \bibnamefont{Tung~Ho}},
      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}}, and
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{year}{2017}, \bibinfo{journal}{Systematic Biology}
      \textbf{\bibinfo{volume}{66}}(\bibinfo{number}{3}), \bibinfo{pages}{299},
      ISSN \bibinfo{issn}{1063-5157}.

    \bibitem[{\citenamefont{Guindon} \emph{et~al.}(2016)\citenamefont{Guindon, Guo,
      and Welch}}]{guindon_demographic_2016}
    \bibinfo{author}{\bibnamefont{Guindon}, \bibfnamefont{S.}},
      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Guo}}, and
      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Welch}},
      \bibinfo{year}{2016}, \bibinfo{journal}{Theoretical Population Biology}
      \textbf{\bibinfo{volume}{111}}, \bibinfo{pages}{43}, ISSN
      \bibinfo{issn}{0040-5809}.

    \bibitem[{\citenamefont{Hallatschek and
      Fisher}(2014)}]{hallatschek_acceleration_2014}
    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
      \bibinfo{author}{\bibfnamefont{D.~S.} \bibnamefont{Fisher}},
      \bibinfo{year}{2014}, \bibinfo{journal}{Proceedings of the National Academy
      of Sciences} \textbf{\bibinfo{volume}{111}}(\bibinfo{number}{46}),
      \bibinfo{pages}{E4911}, \bibinfo{note}{publisher: Proceedings of the National
      Academy of Sciences}.

    \bibitem[{\citenamefont{Hallatschek and Nelson}(2008)}]{hallatschek_gene_2008}
    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
      \bibinfo{year}{2008}, \bibinfo{journal}{Theoretical Population Biology}
      \textbf{\bibinfo{volume}{73}}(\bibinfo{number}{1}), \bibinfo{pages}{158},
      ISSN \bibinfo{issn}{0040-5809}.

    \bibitem[{\citenamefont{Hallatschek and Nelson}(2010)}]{hallatschek_life_2010}
    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
      \bibinfo{year}{2010}, \bibinfo{journal}{Evolution}
      \textbf{\bibinfo{volume}{64}}(\bibinfo{number}{1}), \bibinfo{pages}{193},
      ISSN \bibinfo{issn}{0014-3820}.

    \bibitem[{\citenamefont{Kalkauskas}
      \emph{et~al.}(2021)\citenamefont{Kalkauskas, Perron, Sun, Goldman, Baele,
      Guindon, and Maio}}]{kalkauskas_sampling_2021}
    \bibinfo{author}{\bibnamefont{Kalkauskas}, \bibfnamefont{A.}},
      \bibinfo{author}{\bibfnamefont{U.}~\bibnamefont{Perron}},
      \bibinfo{author}{\bibfnamefont{Y.}~\bibnamefont{Sun}},
      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Goldman}},
      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Guindon}}, and
      \bibinfo{author}{\bibfnamefont{N.~D.} \bibnamefont{Maio}},
      \bibinfo{year}{2021}, \bibinfo{journal}{PLOS Computational Biology}
      \textbf{\bibinfo{volume}{17}}(\bibinfo{number}{1}),
      \bibinfo{pages}{e1008561}, ISSN \bibinfo{issn}{1553-7358},
      \bibinfo{note}{publisher: Public Library of Science}.

    \bibitem[{\citenamefont{Kingman}(1982)}]{kingman_coalescent_1982}
    \bibinfo{author}{\bibnamefont{Kingman}, \bibfnamefont{J.~F.~C.}},
      \bibinfo{year}{1982}, \bibinfo{journal}{Stochastic Processes and their
      Applications} \textbf{\bibinfo{volume}{13}}(\bibinfo{number}{3}),
      \bibinfo{pages}{235}, ISSN \bibinfo{issn}{0304-4149}.

    \bibitem[{\citenamefont{Kolmogorov}
      \emph{et~al.}(1937)\citenamefont{Kolmogorov, Petrovskii, and
      Piscunov}}]{KPP1937}
    \bibinfo{author}{\bibnamefont{Kolmogorov}, \bibfnamefont{A.}},
      \bibinfo{author}{\bibfnamefont{I.}~\bibnamefont{Petrovskii}}, and
      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Piscunov}},
      \bibinfo{year}{1937}, \bibinfo{journal}{Byul. Moskovskogo Gos. Univ.}
      \textbf{\bibinfo{volume}{1}}(\bibinfo{number}{6}), \bibinfo{pages}{1}.

    \bibitem[{\citenamefont{Layan} \emph{et~al.}(2023)\citenamefont{Layan, Müller,
      Dellicour, De~Maio, Bourhy, Cauchemez, and Baele}}]{layan_impact_2023}
    \bibinfo{author}{\bibnamefont{Layan}, \bibfnamefont{M.}},
      \bibinfo{author}{\bibfnamefont{N.~F.} \bibnamefont{Müller}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Dellicour}},
      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{De~Maio}},
      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Bourhy}},
      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Cauchemez}}, and
      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
      \bibinfo{year}{2023}, \bibinfo{journal}{Virus Evolution}
      \textbf{\bibinfo{volume}{9}}(\bibinfo{number}{1}), \bibinfo{pages}{vead010},
      ISSN \bibinfo{issn}{2057-1577}.

    \bibitem[{\citenamefont{Lemey} \emph{et~al.}(2009)\citenamefont{Lemey, Rambaut,
      Drummond, and Suchard}}]{lemey_bayesian_2009}
    \bibinfo{author}{\bibnamefont{Lemey}, \bibfnamefont{P.}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
      \bibinfo{author}{\bibfnamefont{A.~J.} \bibnamefont{Drummond}}, and
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{year}{2009}, \bibinfo{journal}{PLOS Computational Biology}
      \textbf{\bibinfo{volume}{5}}(\bibinfo{number}{9}), \bibinfo{pages}{e1000520},
      ISSN \bibinfo{issn}{1553-7358}, \bibinfo{note}{publisher: Public Library of
      Science}.

    \bibitem[{\citenamefont{Lemey} \emph{et~al.}(2010)\citenamefont{Lemey, Rambaut,
      Welch, and Suchard}}]{lemey_phylogeography_2010}
    \bibinfo{author}{\bibnamefont{Lemey}, \bibfnamefont{P.}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
      \bibinfo{author}{\bibfnamefont{J.~J.} \bibnamefont{Welch}}, and
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{year}{2010}, \bibinfo{journal}{Molecular Biology and Evolution}
      \textbf{\bibinfo{volume}{27}}(\bibinfo{number}{8}), \bibinfo{pages}{1877},
      ISSN \bibinfo{issn}{0737-4038}.

    \bibitem[{\citenamefont{Pybus} \emph{et~al.}(2012)\citenamefont{Pybus, Suchard,
      Lemey, Bernardin, Rambaut, Crawford, Gray, Arinaminpathy, Stramer, Busch, and
      Delwart}}]{pybus_unifying_2012}
    \bibinfo{author}{\bibnamefont{Pybus}, \bibfnamefont{O.~G.}},
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
      \bibinfo{author}{\bibfnamefont{F.~J.} \bibnamefont{Bernardin}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
      \bibinfo{author}{\bibfnamefont{F.~W.} \bibnamefont{Crawford}},
      \bibinfo{author}{\bibfnamefont{R.~R.} \bibnamefont{Gray}},
      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Arinaminpathy}},
      \bibinfo{author}{\bibfnamefont{S.~L.} \bibnamefont{Stramer}},
      \bibinfo{author}{\bibfnamefont{M.~P.} \bibnamefont{Busch}}, and
      \bibinfo{author}{\bibfnamefont{E.~L.} \bibnamefont{Delwart}},
      \bibinfo{year}{2012}, \bibinfo{journal}{Proceedings of the National Academy
      of Sciences} \textbf{\bibinfo{volume}{109}}(\bibinfo{number}{37}),
      \bibinfo{pages}{15066}, \bibinfo{note}{publisher: Proceedings of the National
      Academy of Sciences}.

    \bibitem[{\citenamefont{Raghwani} \emph{et~al.}(2011)\citenamefont{Raghwani,
      Rambaut, Holmes, Hang, Hien, Farrar, Wills, Lennon, Birren, Henn, and
      Simmons}}]{raghwani_endemic_2011}
    \bibinfo{author}{\bibnamefont{Raghwani}, \bibfnamefont{J.}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
      \bibinfo{author}{\bibfnamefont{E.~C.} \bibnamefont{Holmes}},
      \bibinfo{author}{\bibfnamefont{V.~T.} \bibnamefont{Hang}},
      \bibinfo{author}{\bibfnamefont{T.~T.} \bibnamefont{Hien}},
      \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Farrar}},
      \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Wills}},
      \bibinfo{author}{\bibfnamefont{N.~J.} \bibnamefont{Lennon}},
      \bibinfo{author}{\bibfnamefont{B.~W.} \bibnamefont{Birren}},
      \bibinfo{author}{\bibfnamefont{M.~R.} \bibnamefont{Henn}}, and
      \bibinfo{author}{\bibfnamefont{C.~P.} \bibnamefont{Simmons}},
      \bibinfo{year}{2011}, \bibinfo{journal}{PLOS Pathogens}
      \textbf{\bibinfo{volume}{7}}(\bibinfo{number}{6}), \bibinfo{pages}{e1002064},
      ISSN \bibinfo{issn}{1553-7374}, \bibinfo{note}{publisher: Public Library of
      Science}.

    \bibitem[{\citenamefont{Trovão} \emph{et~al.}(2015)\citenamefont{Trovão,
      Suchard, Baele, Gilbert, and Lemey}}]{trovao_bayesian_2015}
    \bibinfo{author}{\bibnamefont{Trovão}, \bibfnamefont{N.~S.}},
      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}}, and
      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
      \bibinfo{year}{2015}, \bibinfo{journal}{Molecular Biology and Evolution}
      \textbf{\bibinfo{volume}{32}}(\bibinfo{number}{12}), \bibinfo{pages}{3264},
      ISSN \bibinfo{issn}{0737-4038}.

    \bibitem[{\citenamefont{Vaughan} \emph{et~al.}(2014)\citenamefont{Vaughan,
      Kühnert, Popinga, Welch, and Drummond}}]{vaughan_efficient_2014}
    \bibinfo{author}{\bibnamefont{Vaughan}, \bibfnamefont{T.~G.}},
      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Kühnert}},
      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Popinga}},
      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Welch}}, and
      \bibinfo{author}{\bibfnamefont{A.~J.} \bibnamefont{Drummond}},
      \bibinfo{year}{2014}, \bibinfo{journal}{Bioinformatics}
      \textbf{\bibinfo{volume}{30}}(\bibinfo{number}{16}), \bibinfo{pages}{2272},
      ISSN \bibinfo{issn}{1367-4803}.

    \bibitem[{\citenamefont{Wilkins and Wakeley}(2002)}]{wilkins_coalescent_2002}
    \bibinfo{author}{\bibnamefont{Wilkins}, \bibfnamefont{J.~F.}}, and
      \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Wakeley}},
      \bibinfo{year}{2002}, \bibinfo{journal}{Genetics}
      \textbf{\bibinfo{volume}{161}}(\bibinfo{number}{2}), \bibinfo{pages}{873},
      ISSN \bibinfo{issn}{0016-6731}.

    \bibitem[{\citenamefont{Yule}(1925)}]{yule_iimathematical_1925}
    \bibinfo{author}{\bibnamefont{Yule}, \bibfnamefont{G.~U.}},
      \bibinfo{year}{1925}, \bibinfo{journal}{Philosophical Transactions of the
      Royal Society of London. Series B, Containing Papers of a Biological
      Character} \textbf{\bibinfo{volume}{213}}(\bibinfo{number}{402-410}),
      \bibinfo{pages}{21}, \bibinfo{note}{publisher: Royal Society}.

    \end{thebibliography}
